---
title: R Access to Twitter's V2 API
author: brshallo
date: '2022-04-04'
categories:
  - rstats
tags: ["`r funspotr::spot_tags(drop_knitr = TRUE)`"]
slug: notes-on-r-access-to-twitter-s-v2-api
toc: no
thumbnail: /2022/04/04/notes-on-r-access-to-twitter-s-v2-api/images/tweet-tweet-unsplash.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


The [rtweet](https://github.com/ropensci/rtweet) package is still the easiest way to GET and POST Twitter data from R. However its developers are currently [working on](https://github.com/ropensci/rtweet/issues/445) adapting it to the new API. V2 comes with a variety of new features. The one I was interested in was being able to GET the users who liked a tweet.

[academictwitteR](https://github.com/cjbarrie/academictwitteR) is probably the most established package that provides a quickstart entry point to the V2 API. However it requires creating an academic account in twitter, i.e. the user must be affiliated with a university. I also stumbled onto [RTwitterV2](https://github.com/MaelKubli/RTwitterV2) and [voson.tcn](https://github.com/vosonlab/voson.tcn) which both also provide quickstarts on the V2 API, but did not explore these.

Instead I followed the tutorial [Getting started with R and v2 of the Twitter API](https://developer.twitter.com/en/docs/tutorials/getting-started-with-r-and-v2-of-the-twitter-api) by Twitter Developer Advocate [Jessica Garson](https://twitter.com/jessicagarson) that uses `{httr}` to interact more directly with the API. I highly recommend reading her tutorial. The code below is mostly just copied from there but changed to provide an example of getting the usernames of those that liked a tweet^[I will warn that it took me an embarrassing amount of time to catch a slight change in the Twitter API between now and when Jessica wrote her post see [note on `expansions`](https://github.com/ropensci/rtweet/issues/445#issuecomment-1088084784).].

**GETting likers of tweet**

Last year, twitter announced [Likes lookup](https://twittercommunity.com/t/announcing-twitter-api-v2-likes-lookup-and-blocks-lookup/154353) in API v2.

Below is a minimal example of how to find every user who liked any tweet, in my case I'll check favoriters of my initial announcement of the [pwiser](https://github.com/brshallo/pwiser/) package. 

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Compute all pairwise ratios, t-tests, distance metrics, or whatever using {pwiser} a new 📦 I&#39;m writing with <a href="https://twitter.com/CarloooMedina?ref_src=twsrc%5Etfw">@CarloooMedina</a> for applying arbitrary functions on pairwise combinations of columns using a `dplyr::across()` style syntax: <a href="https://t.co/DTXElWbuuX">https://t.co/DTXElWbuuX</a> <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a> <a href="https://twitter.com/hashtag/tidyverse?src=hash&amp;ref_src=twsrc%5Etfw">#tidyverse</a> <a href="https://t.co/wrMzX5mOYx">pic.twitter.com/wrMzX5mOYx</a></p>&mdash; Bryan Shalloway (@brshallo) <a href="https://twitter.com/brshallo/status/1394072770661822464?ref_src=twsrc%5Etfw">May 16, 2021</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

First you need to:

1. Apply for a Twitter developer account.
2. Create a project
3. Create an app within that project^[Think you need the app *within* a project, not just a standalone app for access to V2. The **Creating a Twitter app** section of [Authentication with rtweet](https://docs.ropensci.org/rtweet/articles/auth.html#creating-a-twitter-app) is also a helpful tutorial for getting started here. However the documentation there does not indicate the need to create a project first, however `{rtweet}` uses V1, not V2.]

Once you've done this, you'll need to save your BEARER TOKEN somewhere secure, see Hadley's write-up on [Managing secrets](https://cran.r-project.org/web/packages/httr/vignettes/secrets.html) and use your choice of approach. For the write-up below I'll load the secret from the environment^[Project has a local .Renviron file.].

```{r}
library(rjson)
require(httr)
require(jsonlite)
require(dplyr)
library(purrr)

bearer_token <- Sys.getenv("TWITTER_BEARER")
headers <- c(`Authorization` = sprintf('Bearer %s', bearer_token))

tweet_id <- "1394072770661822464"
url_handle <- glue::glue("https://api.twitter.com/2/tweets/{status_id}/liking_users", status_id = tweet_id)

response <- httr::GET(url = url_handle,
                     httr::add_headers(.headers = headers))
                     # query = params)

obj <- httr::content(response, as = "text")
x <- rjson::fromJSON(obj)
```

And then let's pull out the usernames of the people who liked the tweet:

```{r}
x$data %>% 
  purrr::map_chr("username")
```

One thing you may note is that I don't use the `query` argument, I just pull the default parameters -- again see Jessica's posts for a more sophisticated example as well as descriptions of the steps above.
