---
title: Identifying R Functions & Packages in Github Gists
author: Bryan Shalloway
date: '2022-02-07'
categories:
  - data science
  - rstats
tags:
  - dplyr
  - purrr
  - stringr
  - funspotr
  - readr
  - DT
slug: identifying-r-functions-packages-in-your-github-gists
codefolding_show: show
disable_codefolding: false
thumbnail: /2022/02/07/identifying-r-functions-packages-in-your-github-gists/images/filing-cabinets.jpg
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE)
```

This post is a follow-up to my previous post [Identifying R Functions & Packages Used in GitHub Repos](https://www.bryanshalloway.com/2022/01/18/identifying-r-functions-packages-used-in-github-repos/) that introduced [funspotr](https://github.com/brshallo/funspotr).

funspotr can also be applied to gists:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">By functions or packages used...?<a href="https://t.co/kbSLOpQZLF">https://t.co/kbSLOpQZLF</a></p>&mdash; Bryan Shalloway (@brshallo) <a href="https://twitter.com/brshallo/status/1484921548154621953?ref_src=twsrc%5Etfw">January 22, 2022</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

A problem I bumped into was that most of Chelsea's gists don't actually have .R or .Rmd extensions so my approach skipped most of her snippets. I wanted to parse my own gists but ran into a related problem that most of my github gist code snippets are saved as .md files^[As I often use this output type when creating a [reprex](https://github.com/tidyverse/reprex).] so `knitr::purl()` won't work^[`knitr::purl()` is used in functions within funspotr to parse R markdown files.].

In this post I...

1. create a function to extract code chunks from simple .md files
1. parse the functions and packages in my code using [funspotr](https://github.com/brshallo/funspotr)^[In the future I may do a follow-up that passes the parsed functions and packages through a network analysis or some other approach to better visualize the relationships between code snippets.].


# Parsing code

First I used funspotr to get a table with all of my gists.

```{r, inlcude = FALSE}

```


```{r}
library(dplyr)
library(purrr)
library(stringr)
library(funspotr)
```

```{r}
library(dplyr)
library(purrr)
library(stringr)
library(funspotr)

brshallo_gists <- funspotr::github_gists("brshallo")

brshallo_gists
```

## Parsing R files

funspotr is already set-up to parse all the unique functions and packages from R or Rmd files.

```{r, include = FALSE}
# display output was weird here so just copied code and set eval = FALSE
r_gists <- brshallo_gists %>% 
  filter(funspotr:::str_detect_r_rmd(contents))

r_gists_parsed <- funspotr::github_spot_funs(custom_urls = r_gists)

r_gists_unnested <- r_gists_parsed %>% 
  funspotr::unnest_github_results()
```

```{r, eval = FALSE}
r_gists <- brshallo_gists %>% 
  filter(funspotr:::str_detect_r_rmd(contents))

r_gists_parsed <- funspotr::github_spot_funs(custom_urls = r_gists)

r_gists_unnested <- r_gists_parsed %>% 
  funspotr::unnest_github_results()
```

```{r}
r_gists_unnested
```

## Parsing markdown files

To parse my .md files, I wrote a function here `extract_code_md()` that...

* reads in a file
* extracts the text in code chunks^[based on what exists between ticks. Kind of like a less reliable version of `knitr::purl()` but for .md files. Also posted function on [SO question](https://stackoverflow.com/a/71025744/9059865).]
* saves it to a temporary file
* returns the file path of the temporary file



```{r}
subset_even <- function(x) x[!seq_along(x) %% 2]

extract_code_md <- function(file_path){
  
  lines <- readr::read_file(file_path) %>% 
    stringr::str_split("```.*", simplify = TRUE) %>%
    subset_even() %>% 
    stringr::str_flatten("\n## new chunk \n")
  
  file_output <- tempfile(fileext = ".R")
  writeLines(lines, file_output)
  file_output
}
```

I map `extract_code_md()` on all the .md gists and then parse the files using funspotr.

```{r, include = FALSE, eval = FALSE}
#### More detailed notes ####

#### md_gists_local step ####
# I keep the column name `urls` (even though these are now local) becuase that
# is what unnest_github_results() expects even though it can handle urls or
# local filepaths as well (perhaps should change name in funspotr)

#### md_gists_parsed step ####
# an inefficiency here is that github_spot_funs() is also going to resave all of
# these to another local copy of the file... maybe will fix in the future so
# don't have so much redundant file writing but in this case with these small
# text files it just doesn't really matter.

```


```{r, include = FALSE}
# display output was weird here so just copied code and set eval = FALSE
md_gists <- brshallo_gists %>% 
  filter(!funspotr:::str_detect_r_rmd(contents))

md_gists_local <- md_gists %>% 
  # name urls because that's what funspotr::githup_spot_funs() expects
  mutate(urls = map_chr(urls, extract_code_md))

md_gists_parsed <- funspotr::github_spot_funs(custom_urls = md_gists_local)

md_gists_unnested <- md_gists_parsed %>% 
  funspotr::unnest_github_results()
```

```{r, eval = FALSE}
# display output was weird here so just copied code and set eval = FALSE
md_gists <- brshallo_gists %>% 
  filter(!funspotr:::str_detect_r_rmd(contents))

md_gists_local <- md_gists %>% 
# name urls because that's what funspotr::githup_spot_funs() expects
  mutate(urls = map_chr(urls, extract_code_md))

md_gists_parsed <- funspotr::github_spot_funs(custom_urls = md_gists_local)

md_gists_unnested <- md_gists_parsed %>% 
  funspotr::unnest_github_results()
```

```{r}
md_gists_unnested
```

Note that I'm assuming all the code snippets are R code^[Otherwise the R code parsing steps in funspotr will fail.].

## Binding files together

I bind these files together and then arrange them based on the initial order in `brshallo_gists`^[Note that this will just return the *unique* functions in each file, if I want to see every time I used a function I would have passed in `show_each_use = FALSE` to `github_spot_funs()`.].

```{r}
gists_unnested <- bind_rows(
  r_gists_unnested,
  md_gists_unnested
) %>% 
  # got this arranging by a vector trick from SO:
  # https://stackoverflow.com/questions/52216341/how-to-sort-rows-of-a-data-frame-based-on-a-vector-using-dplyr-pipe
  arrange(match(contents, brshallo_gists$contents)) %>% 
  # add back-in links to url's where files are rather than urls column being
  # local paths for .md snippets
  select(-urls) %>% 
  left_join(brshallo_gists, by = "contents")

gists_unnested %>% 
  DT::datatable(rownames = FALSE,
            class = 'cell-border stripe',
            filter = 'top',
            escape = FALSE,
            options = list(pageLength = 20))

```

# Organizing snippets

Perhaps I'll do a follow-up and show some ways the relationships between the resulting parsed code snippets may be visualized in a network or organized in some other way. 

Mentioned in the initial thread, Obsidian seems to be a product that does some things along these lines:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">I’ve found this useful (<a href="https://t.co/OYzwfTltLG">https://t.co/OYzwfTltLG</a>) I is a tool for writing, organizing, linking markdown files.</p>&mdash; John Lee (@Jdlee888) <a href="https://twitter.com/Jdlee888/status/1484332889609170946?ref_src=twsrc%5Etfw">January 21, 2022</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

# Appendix

Interactively save current gists to folder so can read from another file if want to
```{r, eval = FALSE}
post_path <- fs::path_dir(rstudioapi::getSourceEditorContext()$path)

fs::dir_create(post_path, "data")

readr::write_csv(gists_unnested, fs::path(post_path, "data", paste0("brshallo-gists-", format(Sys.Date(), "%Y%m%d"), ".csv")))
```

