---
title: Tidy Pairwise Operations
author: Bryan Shalloway
date: '2020-05-31'
slug: tidy-2-way-column-combinations
categories:
  - r
  - rstats
  - programming
tags:
  - r
  - rstats
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Say you want to map an operation or list of operations across all two-way[^twoway] combinations of a set of variables/columns in a dataframe. For example, you may be doing feature engineering and want to create a set of interaction terms, ratios, etc.^[Though this "throw everything and the kitchen-sink" approach may not always be a good idea.]. You may be interested in computing a summary statistic across all pairwise combinations of a given set of variables[^1]. In some cases there may be an excellent pairwise implementation already available, e.g. R's `cor()` function for computing correlations^[Or the tidymodels implementation `corrr::correlate()` in the [corrr](https://corrr.tidymodels.org/) package.]. In other cases one may not exist, is not easy to use, or is not in a style you prefer. In this post I'll walk through an example[^examples] explaining code and steps for setting-up arbitrary pairwise operations across sets of variables. 

*I'll break my approach down into several steps:*

I. Nest and pivot  
II. Expand combinations  
III. Filter redundancies  
IV. Map function(s)  
V. Reshape, re-pivot & unnest  
VI. Bind back to data  

**Relevant software and style:**

I will primarily be using R's `tidyverse` packages. I make frequent use of lists as columns within dataframes -- if you are new to these, see my previous [talk](https://www.youtube.com/watch?v=gme4Fb9JVjk) and the resources^[In particular, the chapters on "Iteration" and "Many Models" in [R for Data Science](https://r4ds.had.co.nz/iteration.html)] I link to in the description. 

Throughout this post, wherever I write "dataframe" I really mean "tibble" (a dataframe with minor changes to default options and slightly prettier printing behavior). Also note that I am using `dplyr` 0.8.3 rather than the newly released 1.0.0[^dplyr].

[^twoway]: Will focus on two-way example in this post, but could use similar methods to make more generalizable solution across n-way examples. If I were to do this, the code below would change. E.g.

    * to use `pmap*()` operations over `map2*()` operations
    * I'd need to make some functions that make it so I can remove all the places where I have `var` and `var2` type column names hard-coded
    * Alternatively, I might shift approaches and make better use of `combn()`

[^1]: I've done this type of operation in a variety of ways. Sometimes without any really good reason as to why I used one approach or another. It isn't completely clear (at least to me) the recommended way of doing these type of operations within the tidyverse -- hence the diversity of my approaches in the past and deciding to document the typical steps in the approach I take... via writing this post. 

[^examples]: I'll also reference related approaches / small tweaks (putting those materials in the [Appendix]. This is by no means an exhaustive list (e.g. don't have an example with a `for` loop or with a `%do%` operator). The source code of my post on [Ambiguous Absolute Value](https://www.bryanshalloway.com/2020/02/13/fivethirtyeightriddlersolutions-palindrome-debts-and-ambiguous-absolut-value-signs/) signs shows a related but more complex / messy approach on a combinatorics problem.

[^dplyr]: The new `dplyr` 1.0.0. contains new functions that would have been potentially useful for several of these operations. I highly recommend checking these updates out in the various [recent posts](https://www.tidyverse.org/blog/) by Hadley Wickham. Some of the major updates (potentially relevant to the types of operations I'll be discussing in my post):
    
    * new approach for across-column operations (replacing `_at()`, `_if()`, `_all()` variants with `across()` function)
    * brought-back rowwise operations
    * emphasize ability to output tibbles / multiple columns in core `dplyr` verbs... This is something I had only taken advantage of occassionally in the past ([example](https://stackoverflow.com/a/54725732/9059865)), but will look to use more going forward.]

```{r}
library(tidyverse)
```

# Data

I'll use the ames housing dataset across examples.

```{r}
ames <- AmesHousing::make_ames()
```

Specifically, I'll focus on ten numeric columns that, based on a random sample of 1000 rows, show the highest correlation with `Sale_Price`[^technical]. 

[^technical]: For technical reasons, I also converted all integer types to doubles -- was getting integer overflow problems in later operations before changing. [Thread](https://stackoverflow.com/questions/8804779/what-is-integer-overflow-in-r-and-how-can-it-happen) on integer overflow in R. In this post I'm not taking a disciplined approach to feature engineering, however I did standardize the variables so that variable combinations would at least be starting on a similar scale. 

```{r}
set.seed(2020)
ames_cols <- ames %>% 
  select_if(is.numeric) %>% 
  sample_n(1000) %>% 
  corrr::correlate() %>% 
  corrr::focus(Sale_Price) %>% 
  arrange(-abs(Sale_Price)) %>% 
  head(10) %>% 
  pull(rowname)

ames_subset <- select(ames, ames_cols) %>% 
  mutate_all(as.double) %>% 
  # standardize variables
  mutate_all(~(. - mean(.)) / sd(.))
```

# Overview 

## I. Nest and pivot

There are a variety of ways to make lists into columns within a dataframe. In the example below, I use `summarise_all(ames_subset, list)` to create a one row dataframe where each column is a list containing a single element and each individual element corresponds with a numeric vector of length 2930. 

After nesting, I pivot^[Note that this part of the problem is one where I actually find using `tidyr::gather()` easier -- but I've been forcing myself to switch over to using the `pivot_()` functions over `spread()` and `gather()`.] the columns leaving a dataframe with two columns: 

* `var` the variable names
* `vector` a list where each element contains the associated vector

```{r}
df_lists <- ames_subset %>% 
  summarise_all(list) %>% 
  pivot_longer(cols = everything(), 
               names_to = "var", 
               values_to = "vector") %>% 
  print()
```

See [Pivot and then summarise] for a nearly identical approach with just an altered order of steps. Also see [Nested tibbles] for how you could create a list-column of dataframes^[The more common approach.] rather than vectors. 

## II. Expand combinations

I then use `tidyr::nesting()` within `tidyr::expand()` to make all 2-way combinations.
```{r}
df_lists_comb <- expand(df_lists,
                        nesting(var, vector),
                        nesting(var2 = var, vector2 = vector)) %>% 
  print()
```

See [Expand via join] for an alternative approach using the `dplyr::*_join()` operations.

*You could make a strong case that this step should be after [III. Filter redundancies]^[As switching these would be more computationally efficient -- see [When is this approach inappropriate?] for notes related to this.]. However putting it beforehand makes the required code easier to write and to read.*

## III. Filter redundancies

Filter-out redundant columns, sort the rows, better organize the columns.
```{r}
df_lists_comb <- df_lists_comb %>% 
  filter(!(var == var2)) %>% 
  arrange(var, var2) %>% 
  mutate(vars = paste0(var, ".", var2)) %>% 
  select(contains("var"), everything()) %>% 
  print()
```

If your operation of interest is associative^[I.e. has the same output regardless of the order of the variables. E.g. multiplication or addition but not subtraction or division.], apply a filter to remove additional redundant combinations. 
```{r}
c_sort_collapse <- function(...){
  c(...) %>% 
    sort() %>% 
    str_c(collapse = ".")
}

df_lists_comb_as <- df_lists_comb %>% 
  mutate(vars = map2_chr(.x = var, 
                         .y = var2, 
                         .f = c_sort_collapse)) %>%
  distinct(vars, .keep_all = TRUE)
```

## IV. Map function(s)

Each row of your dataframe now contains the relevant combinations of vectors and is ready to have any arbitrary function(s) mapped across them.

**Example with summary statistic^[Function(s) that output vectors of length 1 (or less than length of input vectors).]:**

For example, let's say we want to compute the p-value of the correlation coefficient for each pair[^psych].

[^psych]: Note that the pairwise implementation `psych::corr.test()` could have been used on your original subsetted dataframe, see [stack overflow thread](https://stackoverflow.com/questions/13112238/a-matrix-version-of-cor-test).

```{r}
pairs_cor_pvalues <- df_lists_comb_as %>% 
  mutate(cor_pvalue = map2(vector, vector2, cor.test) %>% map_dbl("p.value"),
         vars = fct_reorder(vars, -cor_pvalue)) %>% 
  arrange(cor_pvalue) %>% 
  print()
```

For fun, let's plot the most significant associations onto a bar graph.
```{r}
pairs_cor_pvalues %>% 
  head(15) %>% 
  mutate(cor_pvalue_nlog = -log(cor_pvalue)) %>% 
  ggplot(aes(x = vars, 
             y = cor_pvalue_nlog, 
             fill = is.infinite(cor_pvalue_nlog) %>% factor(c(T, F))))+
  geom_col()+
  coord_flip()+
  theme_bw()+
  labs(title = "We are confident that garage area and # of garage cars are correlated",
       y = "Negative log of p-value of correlation coefficient",
       x = "Variable combinations",
       fill = "Too high to\nmeaningfully\ndifferentiate:")+
  theme(plot.title.position = "plot")
```

You could use this approach to calculate any pairwise summary statistic. For example, see [gist](https://gist.github.com/brshallo/dc3c1f2f34519ca2a8a68024bc3a22e5) where I calculate the K-S statistic across each combination of a group of distributions.

**Example with transformations^[Function(s) that output vector of length equal to length of input vectors.]:**

Back to the feature engineering example, perhaps we want to calculate the difference and quotient of each combination of our variables.
```{r}
new_features_prep1 <- df_lists_comb %>% 
  mutate(difference = map2(vector, vector2, `-`),
         ratio = map2(vector, vector2, `/`))
```

## V. Reshape, re-pivot & unnest

These steps put our data back into a more traditional form consistent with our starting dataframe/tibble.

**Reshape:**

The goal is to get our dataframe back into a form similar to where it was at the end of [I. Nest and pivot] where we had two columns:

* one with our variable names 
* a second containing a list-column of vectors

*At the end of each line of code in this section is a number corresponding with the explanations below:*

1. if we had done just one operation, this step would not be needed, but we did multiple operations and hence created multiple list-columns (`difference` and `ratio`) which we need to get into a single list-column
2. create new variable name that combines constituent variable names with name of transformation
3. remove old columns

```{r}
new_features_prep2 <- new_features_prep1 %>% 
  pivot_longer(cols = c(difference, ratio)) %>% # 1
  mutate(name_vars = str_c(var, name, var2, sep = ".")) %>% # 2
  select(name_vars, value) %>% # 3
  print()
```

**Re-pivot and unnest:**

Next we simply apply the inverse of those operations performed in [I. Nest and pivot].

```{r}
new_features <- new_features_prep2 %>% 
  pivot_wider(values_from = value,
              names_from = name_vars) %>%
  unnest(cols = everything())
```

The new features will add a good number of columns onto our original dataset^[Did not print this output because cluttered-up page with so many column names.].
```{r}
dim(new_features)
```

## VI. Bind back to data

I then bind the new features back onto the original subsetted dataframe.
```{r}
ames_data_features <- bind_cols(ames_subset, new_features)
```

At which point I could do further exploring, feature engineering, model building, etc.

# Functionalize

I put these steps into a few functions found at [this gist](https://gist.github.com/brshallo/f92a5820030e21cfed8f823a6e1d56e1)^[Steps I - III and V & VI are essentially direct copies of the code above. The approach I took with Step IV may take more effort to follow as it requires understanding a little `rlang` and could likely have been done more simply.]. 

```{r}
devtools::source_gist("https://gist.github.com/brshallo/f92a5820030e21cfed8f823a6e1d56e1")
```

`operations_combinations()` takes in your dataframe, the set of numeric columns to create pairwise combinations from, and a list of functions^[Must have two vectors as input, but do not need to be infix functions.] to apply.

```{r}
ames_data_features_example <- ames %>% 
  keep(is.numeric) %>% 
  mutate_all(as.double) %>% 
  mutate_all(~(. - mean(.)) / sd(.))

ames_data_features_example <- operations_combinations(
  df = ames_data_features_example,
  one_of(ames_cols),
  funs = list("/", "-"),
  funs_names = list("ratio", "difference"),
  associative = FALSE
)
```

To keep things simple, I'm going to remove any columns that contain any NA's or infinite values. I'll also use a random sample of just 1500 rows.
```{r}
features_keep <- ames_data_features_example %>% 
  summarise_all(~sum(is.na(.) | is.infinite(.))) %>% 
  gather() %>% # gather() is an older version of pivot_longer() w/ fewer features
  filter(value == 0) %>% 
  pull(key)

set.seed(1234)
ames_sp_features_subset <- ames_data_features_example %>% 
  select(features_keep) %>% 
  sample_n(1500)
```

Next maybe you are interested in seeing the p-value of the correlation coefficient of each of your features with `Sale_Price`. I'll calculate these across variables and then plot them on a histogram.
```{r}
ames_sp_features_subset %>% 
  select(-Sale_Price) %>% 
  summarise_all(~cor.test(., ames_sp_features_subset$Sale_Price)$p.value) %>% 
  gather() %>% 
  ggplot(aes(x = -log(value)))+
  geom_histogram()+
  scale_x_log10(labels = scales::comma)+
  labs(title = "Distribution of correlations with Sale_Price",
     x = "Negative log of p-value of correlation coefficient")
```

If doing predictive modeling or inference you might fit this type of approach into a `tidymodels` pipeline or other framework. For some brief notes on this see [Interactions example, tidymodels].

# When is this approach innapropriate?

Combinatorial growth is very fast^[Non-technical article discussing combinatorial explosion in context of company user growth targets: [Exponential Growth Isn't Cool. Combinatorial Growth Is.](https://medium.com/@TorBair/exponential-growth-isn-t-cool-combinatorial-growth-is-85a0b1fdb6a5).]. As you increase either the number of variables in your pool or the size of each set, you will quickly bump into computational limitations.

Tidyverse packages are optimized to be computationally efficient. However operations with matrices or other specialized formats^[E.g. [data.table](https://github.com/Rdatatable/data.table) dataframes] are generally faster^[Hence, if you are doing operations across combinations of lots of variables it may not make sense to do the operations directly within dataframes.] than with dataframes/tibbles. If you are running into computational challenges but prefer to stick with a tidyverse aesthetic (which uses dataframes as a cornerstone), you might:

* Use heuristics to reduce the number of variables or operations you need to perform (e.g. take a sample, use a preliminary filter, a step-wise like iteration, etc.)
* Look for packages that abstract the storage and computationally heavy operations away[^2] and then return back an output in a convenient form^[For tidyverse packages, this is often returned into or in the form of a dataframe.]
* Improve the efficiency of your code (e.g. *filter redundancies* before rather than after *expanding combinations*)^[Could make better use of `combn()` function to help.]
* Consider parralelizing
* Use matrices^[Depending on the complexity may just need to brush-up on your linear algebra.]

[^2]: Much (if not most) of the `tidyverse` (and the R programming language generally) is about creating a smooth interface between the analyst/scientist and the back-end complexity of the operations they are performing. Projects like [sparklyr](https://spark.rstudio.com/), [DBI](https://db.rstudio.com/dbi/), [reticulate](https://github.com/rstudio/reticulate), [tidymodels](https://www.tidymodels.org/), and [brms](https://github.com/paul-buerkner/brms) (to name a few) represent cases where this *interface* role of R is most apparent.

For example, you *could* use an approach like the one I walk through to calculate the pairwise correlations between each of your variables. However, the `cor()` function would do this much more efficiently (or you could also use the `corrr` package within the `tidymodels` suite which calls `cor()` in the back-end^[and can also be used to run the operation on databases that may have larger data than you could fit on your computer]).

However, for many operations...

* there may not be an efficient pairwise implementation available / accessible
* the slower computation may not matter or can be mitigated in some way

These situations^[Likely more common for many, if not most, analysts and data scientists.] are where the approach I walked through is most appropriate.

# Appendix

## Interactions example, tidymodels

A good example for creating and evaluating interaction terms^[I.e. multiplying two variables together] is in [The Brute-Force Approach to Identifying Predictive Interactions, Simple Screening](http://www.feat.engineering/complete-enumeration.html#complete-enumeration-simple-screening) section of *Max Kuhn* and *Kjell Johnson's* (free) online book "Feature Engineering and Selection: A Practical Approach for Predictive Models".

The [source code](https://github.com/topepo/FES/blob/master/07_Detecting_Interaction_Effects/7_04_The_Brute-Force_Approach_to_Identifying_Predictive_Interactions/ames_pairwise.R) shows another approach for combining variables.  The author uses...

* `combn()` to create all combinations of variable names which are then... 
* turned into formulas and passed into `recipes::step_interact()`, specifying the new columns to be created^[Created upon the recipe being *baked* or *juiced* -- if you have not checked it out, [recipes](https://github.com/tidymodels/recipes) is AWESOME!]...
* for each interaction term... 
* in each associated model being evaluated 

The example uses a mix of packages and styles and is not a purely tidy approach -- `tidymodels` has also gone through a lot of development since "Feature Engineering and Selection..." was published in 2019[^futurewriteup]. Section 11.2 on [Greedy Search Methods, Simple Filters](http://www.feat.engineering/greedy-simple-filters.html) is also highly relevant.

[^futurewriteup]: Maybe at a future date I'll make a post writing out the example here using the newer approaches now available in `tidymodels`. [Gist](https://gist.github.com/brshallo/674ff06608c1a55fefb8d5dc49896d65) of `combn_ttible()`... starting place for if I ever get to that write-up.

## Expand via join

You can take advantage of join^[Could also have used `right_join()` or `full_join()`.] behavior to create all possible row combinations. In this case, the output will be the same as shown when using `expand()` (except row order will be different).
```{r, eval = FALSE}
left_join(mutate(df_lists, id = 1),
          mutate(df_lists, id = 1) %>% rename_at(vars(-one_of("id")), paste0, "2")) %>%
  select(-id)
```

## Nested tibbles

Creates list of tibbles rather than list of vectors -- typically the first way lists as columns in dataframes is introduced.
```{r, eval = FALSE}
ames_subset %>% 
  pivot_longer(everything(), names_to = "var", values_to = "list") %>% 
  group_by(var) %>% 
  nest()
```

## Pivot and then summarise

(Almost) equivalent to the example in [I. Nest and pivot]. Steps just run in a different order (row order will also be different).

```{r, eval = FALSE}
ames_test %>% 
  pivot_longer(cols = everything(), 
             names_to = "var", 
             values_to = "vector") %>% 
  group_by(var) %>% 
  summarise_all(list)
```

## Session info

```{r}
sessionInfo()
```
